---
title: "Epiclim Liver Metabolomic - F0 adult females"
author: "Baptiste Ducarme"
date: "2025-10-14"
output: 
  html_document:
    code_folding: show       # Montre le code dans le document
    number_sections: true    # Ajout table des matières 
    toc: true                # Table des matiere 
    toc_float: true          # Table des matières flottante
editor_options: 
  chunk_output_type: inline
---

# Introduction

This script was used to perform liver metabolite profiling in adult female zebrafish (15.5–16.5 months post-fertilization) following developmental exposure (2 hpf to 28 dpf) to combined climate stressors, including acidification (pH 6.35) and elevated temperature (32°C), as well as PFOS exposure (nominal concentration: 2 µg/L), using a 2 × 2 × 2 factorial design.

The metabolomic data, consisting of m/z, retention time features and their associated intensities, were first preprocessed in MZmine, including peak detection, alignment, normalization to internal standards, and filtering based on the %RSD of pooled quality control samples. The resulting dataset, used for all downstream analyses, contained both identified and unidentified features across all samples.

To address the few missing values, K-nearest neighbor (KNN) imputation was applied, followed by log2 transformation and Pareto scaling to stabilize variance and approximate normality.

Principal component analysis (PCA) was then performed on the transformed data to visualize the overall sample distribution. Differential analysis was conducted using the limma framework, with experimental condition and exposure included as model factors. Contrasts of interest were defined for each condition versus the control, and MA plots were generated as diagnostic plots. In addition to determining whether each combination treatment elicited a response beyond that predicted by the corresponding individual stressors, pairwise comparisons were performed between each combination treatment and its respective single-stressor treatments.

From the limma output for the condition-versus-control contrasts, the m/z, retention time, t-statistic, and raw p-value were extracted for subsequent pathway analysis. Pathway enrichment analysis was conducted using Mummichog (version 2) in `negative-ionization` mode, with a mass tolerance of 5 ppm and enforcement of primary ions. The top 10% of peaks ranked by p-value were selected as the significance cutoff. The *Danio rerio* (`dre_mtf`) pathway library was used, including only pathway sets containing at least three features. To estimate the null distribution, 100 random permutations were performed. Statistically significant metabolic pathway alterations were defined as those with a Gamma p-value \< 0.05 and Hits sig ≥ 3.

Pipeline inspired by [(Stephen C. Grace and Dane A. Hudson)](https://scispace.com/pdf/processing-and-visualization-of-metabolomics-data-using-r-2f2mbc7w6u.pdf) and [(Jun Sun a Yinglin Xia)](https://doi.org/10.1016/j.gendis.2023.04.018)

## Package

```{r package}
library("readxl") # open data
library("dplyr") # code in pipe
library("ggplot2") # for graph
library("rstatix") # Analyse stat
library("gridExtra") # For graph alignment
library("ggpubr")# for stat link with ggplot
library("lubridate") #for timpoint analysis
library("multcomp")# stat and letter
library('AER')# stat and letter
library("stringr")
library("ggExtra") # PCA ggplot

library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(RColorBrewer) # for a colourful plot
library(ggrepel) # for nice annotations

R.version
```

# Working environment setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Import dataset

```{r}
counts <- read_excel("data/EPICLIM_F0_Liver_Metabolite_Females.xlsx", sheet = "NormData")
count<-as.matrix(counts)
head(counts, 2)
```

## Data treatment

```{r}
# Convert to matrix
counts_matrix <- as.matrix(counts) 

# metabolite matrix with the first column as name
metabolite_names <- counts[[1]] 
counts_matrix <- as.matrix(counts[,-1]) 
counts_matrix <- matrix(as.numeric(counts_matrix), nrow = nrow(counts_matrix))
rownames(counts_matrix) <- metabolite_names
colnames(counts_matrix) <- colnames(counts)[-1] 
```

# Import MetaData

```{r}
metadata<-read_excel("/Users/baptisteducarme/Desktop/EpiCLim-F0-liver-female/Metabolites/EPICLIM_F0_Liver_Metabolite_Females.xlsx",sheet = "MetaData")

head(metadata)
metadata <- metadata %>%rename("Sample"= "name")
metadata <- metadata %>%rename("Fish_nb"= "Fish nb")
colnames(metadata)
```

## MetaData treatment

```{r}
metadata$Sample <- factor(metadata$Sample)
metadata$Condition <- factor(metadata$Condition)
metadata$Exposure <- factor(metadata$Exposure)
metadata$Sex <- factor(metadata$Sex)
metadata$Fish_nb <- factor(metadata$Fish_nb)

library("tibble")
metadata <- column_to_rownames(metadata, var = "Sample")
metadata <- metadata[colnames(counts_matrix), ]

# confirm alignment
all(rownames(metadata) == colnames(counts_matrix))
metadata_Pool <- metadata

# remove pooled samples
metadata <- metadata[metadata$Sex != "Pooled", ]
# remove pooled columns
counts_matrix <- counts_matrix[, rownames(metadata)]

# check dimensions
dim(counts_matrix)

# number of female samples per condition
table(metadata$Condition)
```

## Missing value

[Naniar package](https://naniar.njtierney.com)

[Shadow Matrix](https://cran.r-project.org/web/packages/naniar/vignettes/getting-started-w-naniar.html)

```{r fig.height=8, fig.width=14}
library("naniar")
# diagnostic
mcar_test(as.data.frame(counts_matrix)) 

# missing data heatmap
vis_miss(as.data.frame(counts_matrix))
```

*Fig. 1. Heatmap of missing data across samples.* Black squares indicate missing values, and grey squares represent detected metabolites.

```{r}
library(dplyr)
library(ggplot2)
# logical matrix of missing values (shadow matrix) 
missing_mat <- is.na(counts_matrix)
missing_per_sample <- colMeans(missing_mat) * 100
missing_df <- data.frame( Sample = names(missing_per_sample), MissingPercent = missing_per_sample )
missing_df <- cbind(missing_df, metadata[missing_df$Sample, c("Condition", "Sex")])
# Sort the data
missing_df$Condition <- factor(missing_df$Condition , levels = c("AcidPFOSTemp",
                                                                 "AcidPFOS",
                                                                 "PFOSTemp",
                                                                 "AcidTemp",
                                                                 "PFOS",
                                                                 "Temp",
                                                                 "Acid",
                                                                 "Control"))

ggplot(missing_df, aes(x = MissingPercent, y = Condition, fill = Condition)) +
  geom_boxplot() +
  theme_bw()+
  scale_fill_manual(values = c("Control" = "#4C78A8",
                               "Acid" = "#59A14F",
                               "Temp" = "#E45756",
                               "PFOS" = "#B279A2",
                               "AcidTemp" = "#8C6D1F",
                               "PFOSTemp" = "#AD1457",
                               "AcidPFOS" = "#5B6D5B",
                               "AcidPFOSTemp" = "#7B3F5A")) +
  labs(
    title = "",
    x = "Missing Values (%)",
    y = ""
  ) +
  theme(legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(xlim = c(0, 100))
```

*Fig. 2. Boxplot showing the percentage of missing data across samples.* Colors indicate the experimental condition.

## Inputation (KNN)

[Impute package](https://bioconductor.org/packages/release/bioc/manuals/impute/man/impute.pdf).

Imputation with K-Nearest Neighbors (KNN) estimates missing values in a dataset by considering the values of the closest data points, determined by a distance metric (ussually Euclidean distance). The missing value is then assigned the average of these nearest neighbors’ values, weighted by their proximity ([Sole Galli, 2024](https://www.blog.trainindata.com/knn-imputation-of-missing-values-in-machine-learning/))

**Advantages:**

-   Provides more accurate results than simpler methods like mean or median imputation, especially for complex datasets (Accuracy) ([Piyush Kashyap, 2024](https://medium.com/@piyushkashyap045/handling-missing-values-in-data-a-beginner-guide-to-knn-imputation-30d37cc7a5b7)).

-   Preserve the original variable distribution (preserves the local structure and relationships within the dataset) ([Sole Galli, 2024](https://www.blog.trainindata.com/knn-imputation-of-missing-values-in-machine-learning/)).

-   KNN does not make assumptions about the data distribution, making it flexible for various types of datasets ([Sole Galli, 2024](https://www.blog.trainindata.com/knn-imputation-of-missing-values-in-machine-learning/)).

**Disavantage:**

-   Computationally Expensive: KNN requires distance calculations, which can be slow for large datasets [Piyush Kashyap, 2024](https://medium.com/@piyushkashyap045/handling-missing-values-in-data-a-beginner-guide-to-knn-imputation-30d37cc7a5b7).

-   KNN algorithms do not scale, so their application is usually limited to smaller dataset ([Sole Galli, 2024](https://www.blog.trainindata.com/knn-imputation-of-missing-values-in-machine-learning/)).

-   If too many missing values exist, it struggles to find suitable neighbors ([Taran Kau, 2025](https://medium.com/@tarangds/knn-imputation-the-complete-guide-146f932870a7)).


```{r fig.height=8, fig.width=14}
library("impute")
imputed <- impute.knn(as.matrix(counts_matrix))$data
dim(imputed)

# Missing data heatmap
vis_miss(as.data.frame(imputed))
```

*Fig. 3. Heatmap of missing data across samples, after imputation.* Black squares indicate missing values, and grey squares represent detected metabolites

## Log Transformation

A log transformation helps to remove heteroscedasticity from the data and correct for a skewed data distribution ([Grace & Hudson](http://dx.doi.org/10.5772/65405)).

```{r}
# add a small constant to avoid log(0)
log_counts <- log2(imputed + 1e-9)

dim(log_counts)
```

## Pareto scaling

[Pareto scaling](https://analyticalsciencejournals.onlinelibrary.wiley.com/doi/full/10.1002/cem.3588)

Scaling aims to correct for differences in the magnitude and variance of lipid features by transforming the data relative to a scaling factor. In omics datasets, scaling is applied for several reasons, including reducing differences in feature scale, addressing heteroscedasticity, and improving comparability among subsets of variables with different abundance ranges [(Sun & Xia, 2024)](https://doi.org/10.1016/j.gendis.2023.04.018)..

Pareto scaling provides an intermediate scaling effect between no scaling and unit variance scaling by dividing each mean-centered variable by the square root of its standard deviation [(Sun & Xia, 2024)](https://doi.org/10.1016/j.gendis.2023.04.018).. This approach reduces the influence of highly abundant features while preserving more of the original variance structure compared with unit variance scaling.

In contrast, auto scaling (i.e., mean centering followed by division by the standard deviation) gives equal weight to all variables, which can increase the relative contribution of low-abundance variables that may be more susceptible to measurement error [(Grace & Hudson)](http://dx.doi.org/10.5772/65405)..
```{r}
library("IMIFA")
centred_counts<-pareto_scale(log_counts,centering = TRUE)
```

```{r}
# Log2
df_Log2 <- as.data.frame(log_counts) %>%
  pivot_longer(everything(), names_to = "Sample", values_to = "Intensity") %>%
  mutate(Status = "Before Pareto")

# Pareto
df_Pareto <- as.data.frame(centred_counts) %>%
  pivot_longer(everything(), names_to = "Sample", values_to = "Intensity") %>%
  mutate(Status = "After Pareto")

# combine
df_plot <- bind_rows(df_Log2, df_Pareto)
df_plot$Status <- factor(df_plot$Status, levels = c("Before Pareto", "After Pareto"))

# density plots 
ggplot(df_plot, aes(x = Intensity, fill = Status, color = Status)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~Status, scale="free") +
  labs(title = "",
    x = "Intensity",
    y = "Density") + 
  theme_bw() + 
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20")
```

*Fig. 4. Density plot showing metabolite data transformation and centering across the entire dataset.* The plot includes data before Pareto scaling (already log2-transformed) and after Pareto scaling (centered and scaled).

# PCA

## All

```{r fig.height=8, fig.width=12}
pca_res <- prcomp(t(centred_counts), scale. = FALSE)
var_exp <- round(100 * (pca_res$sdev^2 / sum(pca_res$sdev^2))[1:2], 1)

# PCA dataframe 
pca_df <- data.frame(Sample = colnames(centred_counts),
                      PC1 = pca_res$x[,1], 
                      PC2 = pca_res$x[,2],
                      Condition = metadata$Condition,
                     Exposure = metadata$Exposure)

pca_df$Condition <- factor(pca_df$Condition, levels = c("Control", "Acid", "Temp", "PFOS", "AcidTemp", "PFOSTemp", "AcidPFOS", "AcidPFOSTemp"))


# Distribution by Exposure
PCA_Female_1 <-ggplot(pca_df, aes(x = PC1, y = PC2, color = Exposure)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_color_manual(values = c("expo_2" = "grey", "expo_3" = "orange")) + 
  labs(title = "", x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)")) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "bottom") 

# Distribution by condition
PCA_Female_2 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 2.5, alpha = 0.7) + 
  scale_color_manual(values = c("Control" = "#4C78A8",
                               "Acid" = "#59A14F",
                               "Temp" = "#E45756",
                               "PFOS" = "#B279A2",
                               "AcidTemp" = "#8C6D1F",
                               "PFOSTemp" = "#AD1457",
                               "AcidPFOS" = "#5B6D5B",
                               "AcidPFOSTemp" = "#7B3F5A")) +
  labs(
    title = "",
    x = paste0("PC1 (", var_exp[1], "%)"),
    y = paste0("PC2 (", var_exp[2], "%)")
  ) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        axis.text.x = element_text(hjust = 1, size = 10),
        strip.text.x = element_text(size = 10),
        legend.text = element_text(size = 8))


# Distribution by condition with 95% elipse
PCA_Female_3 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_color_manual(values = c("Control" = "#4C78A8",
                               "Acid" = "#59A14F",
                               "Temp" = "#E45756",
                               "PFOS" = "#B279A2",
                               "AcidTemp" = "#8C6D1F",
                               "PFOSTemp" = "#AD1457",
                               "AcidPFOS" = "#5B6D5B",
                               "AcidPFOSTemp" = "#7B3F5A")) +
  stat_ellipse(level = 0.68, type = "t", linetype = 2) +  # 95% CI ellipse
  labs(
    title = "",
    x = paste0("PC1 (", var_exp[1], "%)"),
    y = paste0("PC2 (", var_exp[2], "%)")
  ) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "bottom")

ggarrange(PCA_Female_1,PCA_Female_2,PCA_Female_3,  ncol = 3, nrow = 2, labels = "AUTO", legend = "bottom")
```
*Fig.5. Principal component analysis (PCA) of log2-transformed and KNN-imputed data.* (A) Colored by experimental exposure (B) Colored by experimental condition (C) inlcuding 95% confidence elipse.

# Limma

[Pipeline limma](https://ucdavis-bioinformatics-training.github.io/2018-June-RNA-Seq-Workshop/thursday/DE.html)

```{r}
library("limma")

# drop unused factor levels
metadata$Sex <- droplevels(metadata$Sex)
metadata$Condition <- droplevels(metadata$Condition)
metadata$Exposure <- droplevels(metadata$Exposure)

# keep only matching samples in scaled_counts
centred_counts <- centred_counts[, rownames(metadata)]

# confirm alignment
stopifnot(all(colnames(centred_counts) == rownames(metadata)))

# Recreate the design matrix
design <- model.matrix(~ 0 + Condition + Exposure, data = metadata)
colnames(design)

fit <- lmFit(centred_counts, design)

# contrast matix
contr.matrix<-makeContrasts(
Acid_Female = ConditionAcid - ConditionControl,
Temp_Female = ConditionTemp - ConditionControl,
PFOS_Female = ConditionPFOS - ConditionControl,
AcidTemp_Female = ConditionAcidTemp - ConditionControl,
AcidPFOS_Female = ConditionAcidPFOS - ConditionControl,
PFOSTemp_Female = ConditionPFOSTemp - ConditionControl,
AcidPFOSTemp_Female = ConditionAcidPFOSTemp - ConditionControl,
levels = design)

# contrast matix
colnames(design)
table(metadata$Condition, metadata$Sex)
colnames(design)

fit <- lmFit(centred_counts, design)
fit2 <- contrasts.fit(fit, contr.matrix)
fit3 <- eBayes(fit2)
```

```{r}
summary(decideTests(fit3))
colnames(fit3$coefficients)
```

## MA plot

```{r fig.height=8, fig.width=14}
par(mfrow = c(4, 4))
plotMA(fit2, coef = "Acid_Female", main = "MA-plot Acid vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "Temp_Female", main = "MA-plot Acid vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "PFOS_Female", main = "MA-plot PFOS vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "AcidTemp_Female", main = "MA-plot AcidTemp vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "PFOSTemp_Female", main = "MA-plot PFOSTemp vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "AcidPFOS_Female", main = "MA-plot AcidPFOS vs Control: Female")+abline(h = 0, col = "red")
plotMA(fit2, coef = "AcidPFOSTemp_Female", main = "MA-plot AcidPFOSTemp vs Control: Female")+abline(h = 0, col = "red")
par(mfrow = c(1, 1))
```

*Fig. 6. MA plots of log2-transformed, KNN-imputed metabolite data.* Values are shown before empirical Bayes adjustment. Each plot represents a contrast between experimental conditions and control. The y-axis (M) shows log2-fold changes, and the x-axis (A) shows average log2 expression across samples. Contrasts include Acid vs Control, Temp vs Control, PFOS vs Control, AcidTemp vs Control, AcidPFOS vs Control, PFOSTemp vs Control, and AcidPFOSTemp vs Control.

## Volcanoplot

### Acid_vs_Control

```{r}
data_results<-topTable(fit3, coef="Acid_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "Acid"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the conpound name

# Volcanoplot
Volcano_1<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "Acid vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5)) +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
Acid_Female_df<- df
DEM_Acid_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-Acid_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "Acid_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### Temp_vs_Control

```{r}
data_results<-topTable(fit3, coef="Temp_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "Temp"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_2<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "Temp vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5)) +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
Temp_Female_df<- df

DEM_Temp_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-Temp_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "Temp_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### PFOS_vs_Control

```{r}
data_results<-topTable(fit3, coef="PFOS_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "PFOS"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_3<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "PFOS vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5))  +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
PFOS_Female_df<- df

DEM_PFOS_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-PFOS_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "PFOS_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### AcidTemp_vs_Control

```{r}
data_results<-topTable(fit3, coef="AcidTemp_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "AcidTemp"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_4<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "AcidTemp vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5)) +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
AcidTemp_Female_df<- df

DEM_AcidTemp_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-AcidTemp_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "AcidTemp_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### AcidPFOS_vs_Control

```{r}
data_results<-topTable(fit3, coef="AcidPFOS_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "AcidPFOS"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_5<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "AcidPFOS vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5))  +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
AcidPFOS_Female_df<- df

DEM_AcidPFOS_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-AcidPFOS_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "AcidPFOS_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### PFOSTemp_vs_Control

```{r}
data_results<-topTable(fit3, coef="PFOSTemp_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "PFOSTemp"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_6<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "PFOSTemp vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5)) +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
PFOSTemp_Female_df<- df

DEM_PFOSTemp_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Data frame preparation
Mummichog_data<-PFOSTemp_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "PFOSTemp_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

### AcidPFOSTemp_vs_Control

```{r}
data_results<-topTable(fit3, coef="AcidPFOSTemp_Female", sort.by="P", number = Inf)
data_results$Compound <- rownames(data_results)
```

```{r}
# Load the ggtext package
library("ggtext")

# Subset your group of interest
df<- data_results[, c("Compound", "logFC","P.Value", "t")]

df$Treatment<- "AcidPFOSTemp"
df$Sex<- "Female"

# Subset your group of interest
colnames(df)[colnames(df) == "P.Value"] <- "pval"
colnames(df)[colnames(df) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
df$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
df$diffexpressed[df$log2fc > 0.6 & df$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
df$diffexpressed[df$log2fc < -0.6 & df$pval < 0.05] <- "DOWN"

# add a colum with IDs but only for the significant protein.
df$label <- ifelse(df$diffexpressed != "NO", df$Compound, NA)
# only identified metabolite
df$label <- ifelse(grepl("^NI@", df$Compound), "", df$label)
df$label <- sub("@.*", "", df$label) # Keep just the compound name

# Volcanoplot
Volcano_7<-ggplot(data = df, aes(x = log2fc, y = -log10(pval), col = diffexpressed)) + 
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') + 
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2, alpha = 0.4) +
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"),
                     labels = c(paste0("Down (", sum(df$diffexpressed == "DOWN"), ")"),
                                paste0("Not significant (",sum(df$diffexpressed == "NO"), ")"),
                                paste0("Up (", sum(df$diffexpressed == "UP"), ")"))) +
  labs(color = "",
       title = "AcidPFOSTemp vs Control : Female",
       x = expression("log"[2]*"(FC)"),
       y = expression("-log"[10]*"p-value")) +
  theme_test() +
  theme(legend.position = "top", 
        plot.title = element_textbox_simple(
      hjust = 0.5, # Center alignment
      halign = 0.5, # Horizontal alignment inside the box
      fill = "grey90", # Background color
      color = "black", # Text color
      box.color = NA, # No border around the background
      padding = margin(5, 5, 5, 5), # Padding inside the title box
      margin = margin(10, 0, 10, 0)))+ # Space around the title box
  coord_cartesian(xlim = c(-5, 5), ylim = c(0, 5)) +
  geom_text_repel(aes(label = label),# Ajouter les étiquettes pour les points significatifs
                  size = 2, # Taille du texte
                  box.padding = 0.5, # Espacement autour du texte
                  max.overlaps = 200)  # Limite les superpositions
```

```{r}
# Subset (usfull for futur analysis)
AcidPFOSTemp_Female_df<- df

DEM_AcidPFOSTemp_Female <- df[df$diffexpressed != "NO", ]
```

```{r}
# Export the DEM in CSV file for Mummichog on metaboAnalyst
# Data frame preparation
Mummichog_data<-AcidPFOSTemp_Female_df
# extract m/z (numbers after "_mz")
Mummichog_data$mz<- sub(".*_mz([0-9.]+)_.*", "\\1", Mummichog_data$Compound)
# extract RT (numbers after "@" and before _mz)
Mummichog_data$RT  <- as.numeric(str_extract(Mummichog_data$Compound, "(?<=@)[0-9.]+"))

head(Mummichog_data)

# restructuration of the data
Mummichog_data<-Mummichog_data[,c("mz","pval", "t", "RT")]

# Change name to fit with MetaboAnalist expecteation
colnames(Mummichog_data)[colnames(Mummichog_data) == "mz"]  <- "m.z"
colnames(Mummichog_data)[colnames(Mummichog_data) == "pval"] <- "p.value"
colnames(Mummichog_data)[colnames(Mummichog_data) == "t"] <- "t.score"
colnames(Mummichog_data)[colnames(Mummichog_data) == "RT"] <- "r.t"

# Change the name of the file !!!!
write.table(
  Mummichog_data,
  "AcidPFOSTemp_Female_Liver_MZ_list.csv",
  sep = ",",           # comma separator
  dec = ".",           # dot for decimals
  quote = FALSE,
  row.names = FALSE)
```

## Up and down plot

```{r}
# For all DEP
combined_df <- rbind(DEM_Acid_Female,
                     DEM_Temp_Female,
                     DEM_PFOS_Female,
                     DEM_AcidTemp_Female,
                     DEM_PFOSTemp_Female,
                     DEM_AcidPFOS_Female,
                     DEM_AcidPFOSTemp_Female)

combined_df<-combined_df[,c("Compound","log2fc","pval","Treatment", "diffexpressed")]



# Summarize the data: count "UP" and "DOWN" for each combination of Treatment and Sex
# Adjust the counts for stacking: DOWN counts are negative
summary_counts <- combined_df %>%
  group_by(Treatment, diffexpressed) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Count = ifelse(diffexpressed == "DOWN", -Count, Count))

factor(summary_counts$diffexpressed, levels = c("UP", "DOWN"))

summary_counts$Treatment <- factor(summary_counts$Treatment, levels = c("Acid", "Temp", "PFOS", "AcidTemp", "PFOSTemp", "AcidPFOS", "AcidPFOSTemp"))

summary_counts$Sex<-"Female"

# Plot the stacked bar chart
CountPlot<-ggplot(summary_counts,
                     aes(x = Treatment,
                         y = Count,
                         fill = factor(diffexpressed, levels = c("UP", "DOWN")))) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Sex) + 
  labs(
    title = "",
    x = "",
    y = "Numbers of DAMs",
    fill = "Expression") +
  geom_text(aes(label = abs(Count)), 
            position = position_stack(vjust = 0.5), 
            size = 4, color = "white") +
  theme_test() +
  scale_fill_manual(values = c("UP" = "#bb0c00", "DOWN" = "#00AFBB")) +
  geom_hline(yintercept = 0, color = "black") +
  theme(legend.position = c(0.85, 0.85), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 10)) +
  guides(fill = guide_legend(title = NULL))



combined_df_female<-combined_df
combined_df_female$Sex<- "Female"
```

```{r fig.height=10, fig.width=12}
ggarrange(Volcano_1, Volcano_2, Volcano_3, Volcano_4, Volcano_5, Volcano_6, Volcano_7,CountPlot,  ncol = 3, nrow = 3, labels = "AUTO")
```

*Fig. 7. Metabolite expression analyses comparing treatment groups to the control, in female.* Panels (A–G) show volcano plots for the contrasts: (A) Acid vs. Control, (B) Temp vs. Control, (C) PFOS vs. Control, (D) AcidTemp vs. Control, (E) AcidPFOS vs. Control, (F) PFOSTemp vs. Control, and (G) AcidPFOSTemp vs. Control. The X-axis represents the log₂ fold change (FC), and the Y-axis represents the −log₁₀(p-value) for all metabolites (known and unknown). The horizontal line indicates the p-value threshold of 0.05 (on a logarithmic scale). Red dots denote upregulated metabolites, blue dots downregulated metabolites, and grey dots unchanged metabolites. Panel (H) shows a count plot summarizing the differentially expressed metabolites (DEMs) across all contrasts.

## UpSet plot (an up grade of the venn plot)

UpSet plot provides an efficient way to visualize intersections of multiple sets compared to the traditional approaches, i.e. the Venn Diagram.
[ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/upset-plot.html) [UpSetR](https://cran.r-project.org/web/packages/UpSetR/vignettes/basic.usage.html)

```{r}
library("UpSetR")

# Metabolite of interest (DEM for each condition)
lt<- list(
  "Acid" = DEM_Acid_Female$Compound,
  "Temp" = DEM_Temp_Female$Compound,
  "PFOS" = DEM_PFOS_Female$Compound,
  "AcidTemp" = DEM_AcidTemp_Female$Compound,
  "PFOSTemp" = DEM_PFOSTemp_Female$Compound,
  "AcidPFOS" = DEM_AcidPFOS_Female$Compound,
  "AcidPFOSTemp" = DEM_AcidPFOSTemp_Female$Compound)

upset(fromList(lt),
  order.by = "freq",       # order intersections by frequency
  keep.order = FALSE,       # keep the order of sets as in the list
  sets = names(lt))   # explicitly define all sets to include
```

*Fig. 8. UpSet plot showing intersections of differentially expressed metabolites (DEMs) across treatment groups.* Horizontal bars indicate the total number of DEMs per group, and vertical bars show the number of DEMs shared among specific combinations of groups. Single dots represent group-specific DEMs, while connected dots indicate shared DEMs


## Combinations vs single_stressor

To determine whether each combination treatment produces a response beyond what any individual stressor alone would predict, pairwise comparisons were made between each combination and its single stressors. These comparisons used the same log2 fold and the same significance thresholds as the vs-Control comparisons (padj \< 0.05, \|log2FC\| ≥ log2(1.5)).

-   A low DELs count in a comparison means the combination have similar profiles to the single stressor.\

-   A high DELs count means the combination is different from the stressor alone.

```{r}
library("limma")
# Single vs combination comparisons
# drop unused factor levels
metadata$Sex <- droplevels(metadata$Sex)
metadata$Condition <- droplevels(metadata$Condition)
metadata$Exposure <- droplevels(metadata$Exposure)

# keep only matching samples in scaled_counts
centred_counts <- centred_counts[, rownames(metadata)]

# confirm alignment
stopifnot(all(colnames(centred_counts) == rownames(metadata)))

# Recreate the design matrix
design <- model.matrix(~ 0 + Condition + Exposure, data = metadata)
colnames(design)

fit <- lmFit(centred_counts, design)

# contrast matix
contr.matrix<-makeContrasts(
  AcidTemp_vs_Acid = ConditionAcidTemp - ConditionAcid, # AcidTemp vs. Acid and Temp
  AcidTemp_vs_Temp =ConditionAcidTemp -  ConditionTemp,
  PFOSTemp_vs_PFOS = ConditionPFOSTemp - ConditionPFOS,  # PFOSTemp vs. PFOS and Temp
  PFOSTemp_vs_Temp = ConditionPFOSTemp - ConditionTemp,
  AcidPFOS_vs_Acid = ConditionAcidPFOS - ConditionAcid,  # AcidPFOS vs. Acid and PFOS  
  AcidPFOS_vs_PFOS = ConditionAcidPFOS - ConditionPFOS,
  AcidPFOSTemp_vs_Acid = ConditionAcidPFOSTemp - ConditionAcid,   # AcidPFOSTemp vs. Acid, Temp and PFOS  
  AcidPFOSTemp_vs_PFOS = ConditionAcidPFOSTemp - ConditionPFOS,
  AcidPFOSTemp_vs_Temp = ConditionAcidPFOSTemp - ConditionTemp,
  AcidPFOSTemp_vs_AcidTemp = ConditionAcidPFOSTemp - ConditionAcidTemp, # AcidPFOSTemp vs.Secondary stressor combinaison  
  AcidPFOSTemp_vs_PFOSTemp = ConditionAcidPFOSTemp - ConditionPFOSTemp,
  AcidPFOSTemp_vs_AcidPFOS= ConditionAcidPFOSTemp - ConditionAcidPFOS,
  levels = design)

# contrast matix
colnames(design)

fit <- lmFit(centred_counts, design)
fit2 <- contrasts.fit(fit, contr.matrix)
fit3 <- eBayes(fit2)

summary(decideTests(fit3))
```

```{r}
# Get limma result
results_AcidTemp_vs_Acid <-topTable(fit3, coef="AcidTemp_vs_Acid", sort.by="P", number = Inf)
results_AcidTemp_vs_Temp <- topTable(fit3, coef="AcidTemp_vs_Temp", sort.by="P", number = Inf)
results_PFOSTemp_vs_PFOS <- topTable(fit3, coef="PFOSTemp_vs_PFOS", sort.by="P", number = Inf)
results_PFOSTemp_vs_Temp <- topTable(fit3, coef="PFOSTemp_vs_Temp", sort.by="P", number = Inf)
results_AcidPFOS_vs_Acid <- topTable(fit3, coef="AcidPFOS_vs_Acid", sort.by="P", number = Inf)
results_AcidPFOS_vs_PFOS <- topTable(fit3, coef="AcidPFOS_vs_PFOS", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_Acid <- topTable(fit3, coef="AcidPFOSTemp_vs_Acid", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_PFOS <- topTable(fit3, coef="AcidPFOSTemp_vs_PFOS", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_Temp <- topTable(fit3, coef="AcidPFOSTemp_vs_Temp", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_AcidTemp <- topTable(fit3, coef="AcidPFOSTemp_vs_AcidTemp", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_PFOSTemp<- topTable(fit3, coef="AcidPFOSTemp_vs_PFOSTemp", sort.by="P", number = Inf)
results_AcidPFOSTemp_vs_AcidPFOS <- topTable(fit3, coef="AcidPFOSTemp_vs_AcidPFOS", sort.by="P", number = Inf)

# Annotate result data
results_AcidTemp_vs_Acid$Treatment <- "AcidTemp_vs_Acid"
results_AcidTemp_vs_Temp$Treatment <- "AcidTemp_vs_Temp"
results_PFOSTemp_vs_PFOS$Treatment <- "PFOSTemp_vs_PFOS"
results_PFOSTemp_vs_Temp$Treatment <- "PFOSTemp_vs_Temp"
results_AcidPFOS_vs_Acid$Treatment <- "AcidPFOS_vs_Acid"
results_AcidPFOS_vs_PFOS$Treatment <- "AcidPFOS_vs_PFOS"
results_AcidPFOSTemp_vs_Acid$Treatment <- "AcidPFOSTemp_vs_Acid"
results_AcidPFOSTemp_vs_PFOS$Treatment <- "AcidPFOSTemp_vs_PFOS"
results_AcidPFOSTemp_vs_Temp$Treatment <- "AcidPFOSTemp_vs_Temp"
results_AcidPFOSTemp_vs_AcidTemp$Treatment <-"AcidPFOSTemp_vs_AcidTemp"
results_AcidPFOSTemp_vs_PFOSTemp$Treatment<- "AcidPFOSTemp_vs_PFOSTemp"
results_AcidPFOSTemp_vs_AcidPFOS$Treatment <- "AcidPFOSTemp_vs_AcidPFOS"

# all pairwise combination result 
results_Pairwise <- rbind(results_AcidTemp_vs_Acid,
                     results_AcidTemp_vs_Temp,
                     results_PFOSTemp_vs_PFOS,
                     results_PFOSTemp_vs_Temp,
                     results_AcidPFOS_vs_Acid,
                     results_AcidPFOS_vs_PFOS,
                     results_AcidPFOSTemp_vs_Acid,
                     results_AcidPFOSTemp_vs_PFOS,
                     results_AcidPFOSTemp_vs_Temp,
                     results_AcidPFOSTemp_vs_AcidTemp,
                     results_AcidPFOSTemp_vs_PFOSTemp,
                     results_AcidPFOSTemp_vs_AcidPFOS)

results_Pairwise$Sex <- "Female"
results_Pairwise$Compound <- rownames(results_Pairwise)

# Subset your group of interest
results_Pairwise<- results_Pairwise[, c("Compound", "logFC","P.Value", "t", "Treatment", "Sex")]

# rename
colnames(results_Pairwise)[colnames(results_Pairwise) == "P.Value"] <- "pval"
colnames(results_Pairwise)[colnames(results_Pairwise) == "logFC"] <- "log2fc"

# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
results_Pairwise$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP"
results_Pairwise$diffexpressed[results_Pairwise$log2fc > 0.6 & results_Pairwise$pval < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
results_Pairwise$diffexpressed[results_Pairwise$log2fc < -0.6 & results_Pairwise$pval < 0.05] <- "DOWN"

# Subset differential feature
DEM_results_Pairwise <- results_Pairwise[results_Pairwise$diffexpressed != "NO", ]

DEM_results_Pairwise %>%
  group_by(Treatment,diffexpressed) %>%
  summarise(Count = n(), .groups = "drop")
```

-   A low DEG count in a comparison means the two conditions have similar transcriptional profiles = the combination is similar to the single stressor.\
-   A high DEG count means the combination is different from the stressor alone.

# Enrich Pathway


`/!\` The input table provided to Mummichog included the complete feature list, necessary to model the background distribution, with m/z values and retention times to enhance metabolite annotation accuracy ([Mummichog](https://www.metaboanalyst.ca/Secure/mummichog/LibraryView.xhtml)).

[Mummichog](https://www.metaboanalyst.ca/Secure/mummichog/LibraryView.xhtml)

[Mummichog R](https://www.metaboanalyst.ca/resources/vignettes/Functional_Analysis_global_metabolomics.html)

**Useful links:**

-   Comprehensive investigation of pathway enrichment methods for functional interpretation of LC–MS global metabolomics data [(Lu et al, 2022)](https://academic.oup.com/bib/article/24/1/bbac553/6960976).

-   Comparison of popular enrichment methods for untargeted in vitro metabolomics [(Schermer et al, 2025)](https://doi.org/10.1007/s11306-025-02309-0).

-   Environmental chemicals and endogenous metabolites in bile of USA and Norway patients with primary sclerosing cholangitis [(Grant et al 2023)](https://doi.org/10.1093/exposome/osac011).

-   Global metabolomic alterations associated with endocrine-disrupting chemicals among pregnant individuals and newborns [(Puvvula et al, 2025)](https://doi.org/10.1007/s11306-024-02219-7)


```{r}
Data_Pathway <- read_csv2("EPICLIM_F0_Liver_Pathway_Metabolite_Female.csv")
Data_Pathway <- Data_Pathway[Data_Pathway$P_Gamma < 0.05 & Data_Pathway$Hits_sig >= 3,]
```

```{r}
pathway_groups <-list("Glycan biosynthesis and metabolism" = c(
  "Sialic acid metabolism",
  "Aminosugars metabolism",
  "Keratan sulfate biosynthesis",
  "Glycosphingolipid biosynthesis - globoseries",
  "Glycosphingolipid biosynthesis - ganglioseries",
  "Glycosphingolipid biosynthesis - lactoseries",
  "Glycosphingolipid biosynthesis - neolactoseries",
  "Blood Group Biosynthesis",
  "O-Glycan biosynthesis",
  "Glycosphingolipid metabolism",
  "N-Glycan biosynthesis",
  "N-Glycan Degradation",
  "Keratan sulfate degradation",
  "Chondroitin sulfate degradation",
  "Heparan sulfate degradation",
  "Phosphatidylinositol phosphate metabolism",
  "Glycosylphosphatidylinositol(GPI)-anchor biosynthesis",
  "Hyaluronan Metabolism",
  "Nucleotide Sugar Metabolism"), 
  "Carbohydrate metabolism" = c( "Pyruvate Metabolism",
    "Glycolysis and Gluconeogenesis",
    "Pentose phosphate pathway",
    "Propanoate metabolism",
    "Glyoxylate and Dicarboxylate Metabolism",
    "Fructose and mannose metabolism",
    "Galactose metabolism",
    "Hexose phosphorylation",
    "Starch and Sucrose Metabolism",
    "Pentose and Glucuronate Interconversions", 
    "Butanoate metabolism",
    "TCA cycle", 
    "C5-Branched dibasic acid metabolism"), 
  "Nucleotide" = c(
    "Purine metabolism",
    "Pyrimidine metabolism"),
  "Amino acid" = c(
    "Glycine- serine- alanine and threonine metabolism", 
    "Valine- leucine and isoleucine degradation", 
    "Tyrosine metabolism", 
    "Methionine and cysteine metabolism",
    "Tryptophan metabolism",
    "Lysine metabolism",
    "Alanine and Aspartate Metabolism",
    "Aspartate and asparagine metabolism",
    "Arginine and Proline Metabolism",
    "Histidine metabolism",
    "Selenoamino acid metabolism", 
    "Beta-Alanine metabolism",
    "Glutathione Metabolism",
    "Glutamate metabolism"),
  "Cofactors and vitamins" = c(
    "Vitamin A (retinol) metabolism",
    "Vitamin B1 (thiamin) metabolism",
    "Vitamin B2 (riboflavin) metabolism",
    "Vitamin B3 (nicotinate and nicotinamide) metabolism",
    "Vitamin B5 - CoA biosynthesis from pantothenate", 
    "Vitamin B6 (pyridoxine) metabolism","Vitamin B9 (folate) metabolism",
    "Ascorbate (Vitamin C) and Aldarate Metabolism",
    "Biopterin metabolism",
    "Vitamin E metabolism",
    "Vitamin H (biotin) metabolism",
    "Ubiquinone Biosynthesis",
    "CoA Catabolism",
    "Porphyrin metabolism"),
  "Lipid metabolism" = c(
    "Mono-unsaturated fatty acid beta-oxidation",
    "De novo fatty acid biosynthesis",
    "Polyunsaturated fatty acid biosynthesis",
    "Fatty Acid Metabolism",
    "Glycerophospholipid metabolism",
    "Fatty acid activation",
    "Arachidonic acid metabolism",
    "Saturated fatty acids beta-oxidation",
    "Di-unsaturated fatty acid beta-oxidation",
    "Squalene and cholesterol biosynthesis",
    "Omega-3 fatty acid metabolism",
    "Carnitine shuttle",
    "Linoleate metabolism", 
    "Bile acid biosynthesis", 
    "Putative anti-Inflammatory metabolites formation from EPA", 
    "C21-steroid hormone biosynthesis and metabolism",
    "Omega-6 fatty acid metabolism", 
    "Prostaglandin formation from arachidonate",
    "Prostaglandin formation from dihomo gama-linoleic acid", 
    "D4&E4-neuroprostanes formation",
    "Androgen and estrogen biosynthesis and metabolism"),
  "Other" = c(
    "Alkaloid biosynthesis II",
    "Leukotriene metabolism",
    "Xenobiotics metabolism", 
    "Urea cycle/amino group metabolism",
    "Electron transport chain",
    "Proteoglycan biosynthesis"))

# Create a named vector for easy lookup
pathway_to_function <- unlist(lapply(names(pathway_groups), function(name) {
  setNames(rep(name, length(pathway_groups[[name]])), pathway_groups[[name]])
}))

# Add BiologicalFunction column to your data frame
Data_Pathway <- Data_Pathway %>%
  mutate(BiologicalFunction = pathway_to_function[as.character(Pathway)]) %>%
  # For pathways not in the list, assign "Unclassified"
  mutate(BiologicalFunction = ifelse(is.na(BiologicalFunction), "Unclassified", BiologicalFunction))

# remove unadapted pathway
Data_Pathway <- Data_Pathway[!(Data_Pathway$Pathway %in% c(
  "Carbon fixation",
  "Alkaloid biosynthesis II",
  "Starch and Sucrose Metabolism",
  "Glyoxylate and Dicarboxylate Metabolism", 
  "Nitrogen metabolism",
  "Parathio degradation"
)), ]
```

```{r fig.height=10, fig.width=10}
Data_Pathway$Condition <- factor(Data_Pathway$Condition, levels = c("Acid", "Temp", "PFOS","AcidTemp", "PFOSTemp","AcidPFOS", "AcidPFOSTemp"))

pathway_order <- Data_Pathway %>%
  distinct(Pathway, BiologicalFunction) %>%
  arrange(BiologicalFunction, Pathway) %>%
  pull(Pathway)

Data_Pathway <- Data_Pathway %>%
  mutate(
    Pathway = factor(Pathway, levels = pathway_order),
    Condition = factor(Condition, levels = c("Acid", "Temp", "PFOS", "AcidTemp", "PFOSTemp", "AcidPFOS", "AcidPFOSTemp")))

Plot_Pathway_female<-ggplot(Data_Pathway) +
  aes(x = Condition, 
    y = Pathway,
    fill = P_Gamma,           # use fill for the inside color
    size = EnrichFactor) +
  facet_grid(BiologicalFunction ~ Sex, scales = "free_y", space = "free_y",
             labeller = labeller(BiologicalFunction = function(x) {
               sapply(x, function(y) paste(strwrap(y, width = 10), collapse = "\n"))
             })) +
  geom_point(shape = 21, color = "grey10", stroke = 0.3, alpha = 0.8) +  # thin black outline
  labs(
    title = "",
    x = "",
    y = "",
    fill = "P-Value", 
    size = "Enrichment \nfactor"
  ) +
  scale_fill_distiller(palette = "Oranges") +
   theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), 
        axis.text = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(size = 10, angle = 0)) + 
  scale_x_discrete(drop = FALSE) +   # print the condition without affected pathway
  scale_y_discrete(drop = TRUE)

Plot_Pathway_female
```

*Fig. 9. Pathway enrichment analysis of liver metabolomics using the Mummichog algorithm.* Significantly enriched pathways (Gamma p-value < 0.05 and Hits sig ≥3) were identified based on the top 10% of metabolite features with the lowest raw p-values. The enrichment factor represents the ratio between the observed number of significant hits and the expected number of hits per pathway.


## Upsetplot.

UpSet plot provides an efficient way to visualize intersections of multiple sets compared to the traditional approaches, i.e. the Venn Diagram.
[ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/upset-plot.html) [UpSetR](https://cran.r-project.org/web/packages/UpSetR/vignettes/basic.usage.html)

```{r}
library("UpSetR")
library("grid")

# Metabolite of interest (DEM for each condition)
lt<- list(
  "Acid" = Data_Pathway[Data_Pathway$Condition == "Acid",]$Pathway,
  "Temp" = Data_Pathway[Data_Pathway$Condition == "Temp",]$Pathway,
  "PFOS" = Data_Pathway[Data_Pathway$Condition == "PFOS",]$Pathway,
  "AcidTemp" = Data_Pathway[Data_Pathway$Condition == "AcidTemp",]$Pathway,
  "PFOSTemp" = Data_Pathway[Data_Pathway$Condition == "PFOSTemp",]$Pathway,
  "AcidPFOS" = Data_Pathway[Data_Pathway$Condition == "AcidPFOS",]$Pathway,
  "AcidPFOSTemp" = Data_Pathway[Data_Pathway$Condition == "AcidPFOSTemp",]$Pathway)

upset_obj <- UpSetR::upset(
  fromList(lt),
  order.by = "freq",
  sets = names(lt),
  decreasing = TRUE,
  mb.ratio = c(0.65, 0.35),
  text.scale = 1.3,
  point.size = 2,
  line.size = 1,
  mainbar.y.label = "Number of shared pathways",
  sets.x.label = "Pathways per condition",
  scale.sets = "identity"
)

upset_grob <- grid.grabExpr(print(upset_obj))
```
*Fig. 10. UpSet plot showing the intersections of enriched pathways across treatment groups.* Horizontal bars indicate the total number of enriched pathways in each group, whereas vertical bars indicate the number of enriched pathways shared among specific combinations of groups. Single dots represent pathways uniquely enriched in a single group, while connected dots indicate pathways shared among multiple groups.

## Session info
```{r session_info}
sessionInfo()
```
