---
title: "Epiclim Morphology analysis (Length, weight, K, HSI) - F0 adult females"
author: "Manon"
date: "2026-04-21"
output: html_document
---
## Morphology analysis - Introduction
This script was used to perform morphological analyses on adult female zebrafish 
(15.5-16.5 months post-fertilization) following developmental exposure 
(2 hpf to 28 dpf) to combined climate stressors (acidification at pH 6.35, 
elevated temperature at 32°C) and PFOS (nominal 2 µg/L) in a 2×2×2 
factorial design.
Data include fish from two independent developmental exposures (exposure 2 
and exposure 3). A preliminary exposure (exposure 1) was excluded due to 
limited sample size.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Load required packages
```{r load-libraries}
library(MASS)      # for rlm
library(readxl)    # open data
library(dplyr)     # code in pipe
library(ggplot2)   # for graph
library(rstatix)   # stats
library(gridExtra) # graph alignment
library(ggpubr)    # stat link with ggplot
library(multcomp)  # stat and letters
library(emmeans)   # estimated marginal means
library(sfsmisc)   # f.robftest for rlm
library(patchwork)
library(car)       # for Type 2 Anova
```

## Working environment setup
```{r setwd}
# Set your working directory to the location of this Rmd file
# setwd("path/to/EpiCLim-F0-liver/morphology")
```

## Import dataset
```{r data_import}
Data <- read_excel(
  normalizePath("Sampling F0-metadata_females.xlsx"),
  sheet = "All", col_names = TRUE
)
summary(Data)
```
## Data treatment
Remove "_" in some column values.
```{r data_treatment}
Data$Exposure <- gsub("_", "", Data$Exposure)

# Sample sizes summary
Data %>%
  group_by(Condition) %>%
  summarise(n = n(), .groups = "drop") %>%
  print()
```

```{r data_setup}
Data$Fish_nb <- as.factor(Data$Fish_nb)
Data$Condition <- as.factor(Data$Condition)
Data$Exposure <- as.factor(Data$Exposure)
summary(Data)

#Check missing values
morph_vars <- c("Length_thins", "Length", "Weight_g", "BSI", "HSI", "GSI", "K")
print(colSums(is.na(Data[, morph_vars])))

#Color code for conditions (used across all plots)
myColors <- c(
  "Control" = "#4C78A8", "Acid" = "#59A14F", "Temp" = "#E45756",
  "PFOS" = "#B279A2", "AcidTemp" = "#8C6D1F", "PFOSTemp" = "#AD1457",
  "AcidPFOS" = "#5B6D5B", "AcidPFOSTemp" = "#7B3F5A")
```

## PCA
```{r PCA}
#Set reference levels (Control as reference for Condition, Expo 2 as reference)
Data$Condition <- factor(Data$Condition,
  levels = c("Control", "Acid", "Temp", "PFOS",
             "AcidTemp", "AcidPFOS", "PFOSTemp", "AcidPFOSTemp"))
Data$Exposure <- relevel(factor(Data$Exposure), ref = "2")

#Check sample sizes per condition
table(Data$Condition, Data$Exposure)

#PCA on complete lines (no missing values) using raw measurements only
#(indices like K, BSI are derived from these, including them would double-count variance)
morph_metrics <- c("Length", "Weight_mg", "Weight_Liver_g")
Data_pca <- Data[complete.cases(Data[, morph_metrics]), ]
cat("Fish used:", nrow(Data_pca), "out of", nrow(Data), "\n")

pca <- prcomp(Data_pca[, morph_metrics], scale. = TRUE)
var_exp <- round(summary(pca)$importance[2, 1:2] * 100, 1)
summary(pca)

pca_df <- data.frame(
  PC1 = pca$x[, 1], PC2 = pca$x[, 2],
  Condition = Data_pca$Condition,
  Exposure = Data_pca$Exposure)

# By Exposure
p1 <- ggplot(pca_df, aes(PC1, PC2, color = Exposure)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.68) +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       title = "By Exposure") +
  theme_classic()

# By Condition
p2 <- ggplot(pca_df, aes(PC1, PC2, color = Condition)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.68) +
  scale_color_manual(values = myColors) +
  labs(x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       title = "By Condition") +
  theme_classic()

grid.arrange(p1, p2)

combined_PCA <- (p1/p2) +
  plot_annotation(tag_levels = "A")

combined_PCA

ggsave("PCA_females.png", plot = combined_PCA,
       width = 7, height = 5, units = "in", dpi = 600)
```
PCA shows no strict batch effect. Expo 2 and 3 overlap.

## Log2 transform data
```{r log2_transfo}
Data$log2_Length_thins <- log2(Data$Length_thins)
Data$log2_Length <- log2(Data$Length)
Data$log2_Weight <- log2(Data$Weight_g)
Data$log2_Weight_mg <- log2(Data$Weight_mg)
Data$log2_BSI <- log2(Data$BSI)
Data$log2_HSI <- log2(Data$HSI)
Data$log2_GSI <- log2(Data$GSI)
Data$log2_K <- log2(Data$K)

#Check the distribution
par(mfrow = c(3, 2))
hist(Data$HSI, main = "HSI - raw", breaks = 20)
hist(log2(Data$HSI), main = "HSI - log2", breaks = 20)
hist(Data$BSI, main = "BSI - raw", breaks = 20)
hist(log2(Data$BSI), main = "BSI - log2", breaks = 20)
hist(Data$K, main = "K - raw", breaks = 20)
hist(log2(Data$K), main = "K - log2", breaks = 20)

par(mfrow = c(3, 2))
hist(Data$Length_thins, main = "Length thins - raw", breaks = 20)
hist(log2(Data$Length_thins), main = "Length thins - log2", breaks = 20)
hist(Data$Length, main = "Length - raw", breaks = 20)
hist(log2(Data$Length), main = "Length - log2", breaks = 20)
hist(Data$Weight_g, main = "Weight - raw", breaks = 20)
hist(log2(Data$Weight_g), main = "Weight - log2", breaks = 20)
```
## Length analysis
```{r length}
#Diagnostic on standard lm first
model_length <- lm(log2_Length ~ Condition + Exposure, data = Data)
par(mfrow = c(2, 2))
plot(model_length, main = "Length")
#residuals vs fitted = random scatter around zero --> good
#QQ plot: some fish differ strongly from the line --> outliers --> switch to rlm

#Robust regression
model_length_robust <- rlm(log2_Length ~ Condition + Exposure, data = Data)

#Omnibus F-test on Condition
f.robftest(model_length_robust, var = c("ConditionAcid", "ConditionTemp", "ConditionPFOS",
                                        "ConditionAcidTemp", "ConditionAcidPFOS",
                                        "ConditionPFOSTemp", "ConditionAcidPFOSTemp"))
f.robftest(model_length_robust, var = "Exposure3")
#Condition effect present. Exposure doesn't matter --> we can keep our 2 batches as technical replicates

#Dunnett: each treatment vs Control
emm <- emmeans(model_length_robust, ~ Condition)
contrast(emm, method = "trt.vs.ctrl", ref = "Control")

#All pairwise comparisons (BH-adjusted, 28 tests)
pairs(emm, adjust = "BH")

#Compact letter display for the plot
cld_df <- as.data.frame(cld(emm, Letters = letters, adjust = "BH"))
cld_df$.group <- trimws(cld_df$.group)

#Sample sizes
n_df <- Data %>%
  filter(!is.na(log2_Length)) %>%
  group_by(Condition) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n))

#Plot
length_plot_females <- ggplot(Data, aes(x = Condition, y = log2_Length, fill = Condition)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_fill_manual(values = myColors) +
  geom_text(data = cld_df, aes(x = Condition, y = 1.55, label = .group),
            size = 5, inherit.aes = FALSE) +
  geom_text(data = n_df, aes(x = Condition, y = 0.8, label = label),
            size = 4, fontface = "italic", inherit.aes = FALSE) +
  labs(y = "Log2 Standard body length (cm)", x = "", title = "Length") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.text.y = element_text(size = 11),
        axis.title.y = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"))
length_plot_females

ggsave("length_plot.png", plot = length_plot_females,
       width = 7, height = 5, units = "in", dpi = 600)
```
## Weight analysis
```{r weight}
#Diagnostic on standard lm
model_Weight <- lm(log2_Weight_mg ~ Condition + Exposure, data = Data)
par(mfrow = c(2, 2))
plot(model_Weight, main = "Weight")
#residuals vs fitted = random scatter around zero --> good
#QQ plot ok --> keep lm

anova(model_Weight)
#Exposure effect probably due to age (expo 2 fish are older --> bigger)
#Condition effect also present

#Dunnett: each treatment vs Control
emm <- emmeans(model_Weight, ~ Condition)
contrast(emm, method = "trt.vs.ctrl", ref = "Control")

#All pairwise comparisons
pairs(emm, adjust = "BH")

#Compact letter display
cld_df <- as.data.frame(cld(emm, Letters = letters, adjust = "BH"))
cld_df$.group <- trimws(cld_df$.group)

#Sample sizes
n_df <- Data %>%
  filter(!is.na(log2_Weight_mg)) %>%
  group_by(Condition) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n))

#Plot
weight_plot_females <- ggplot(Data, aes(x = Condition, y = log2_Weight_mg, fill = Condition)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_fill_manual(values = myColors) +
  geom_text(data = cld_df, aes(x = Condition, y = 9.5, label = .group),
            size = 5, inherit.aes = FALSE) +
  geom_text(data = n_df, aes(x = Condition, y = 6.8, label = label),
            size = 4, fontface = "italic", inherit.aes = FALSE) +
  labs(y = "Log2 Weight (mg)", x = "", title = "Weight") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.text.y = element_text(size = 11),
        axis.title.y = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"))
weight_plot_females

ggsave("weight_plot.png", plot = weight_plot_females,
       width = 7, height = 5, units = "in", dpi = 600)
```
## HSI analysis
```{r HSI}
#Diagnostic on standard lm
model_HSI <- lm(log2_HSI ~ Condition + Exposure, data = Data)
par(mfrow = c(2, 2))
plot(model_HSI, main = "HSI")
#residuals vs fitted = random scatter around zero --> good
#QQ plot some outliers --> rlm

cooksd <- cooks.distance(model_HSI)
threshold <- 4 / (nrow(Data) - length(coef(model_HSI)))
outliers <- which(cooksd > threshold)
outliers

#Robust regression
model_HSI_robust <- rlm(log2_HSI ~ Condition + Exposure, data = Data)

#Omnibus F-test
f.robftest(model_HSI_robust, var = c("ConditionAcid", "ConditionTemp", "ConditionPFOS",
                                     "ConditionAcidTemp", "ConditionAcidPFOS",
                                     "ConditionPFOSTemp", "ConditionAcidPFOSTemp"))
f.robftest(model_HSI_robust, var = "Exposure3")
#No effects whatsoever.

#Sensitivity check: ANOVA without extreme outliers (exploratory)
Data_no_out <- Data[!(rownames(Data) %in% outliers), ]
model_HSI_aov <- aov(log2_HSI ~ Condition + Exposure, data = Data_no_out)
summary(model_HSI_aov)
#still no effects

#Dunnett: each treatment vs Control
emm <- emmeans(model_HSI_robust, ~ Condition)
contrast(emm, method = "trt.vs.ctrl", ref = "Control")

#All pairwise comparisons (BH-adjusted)
pairs(emm, adjust = "BH")

#Compact letter display
cld_df <- as.data.frame(cld(emm, Letters = letters, adjust = "BH"))
cld_df$.group <- trimws(cld_df$.group)

#Sample sizes
n_df <- Data %>%
  filter(!is.na(log2_HSI)) %>%
  group_by(Condition) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n))

#Plot
plot_HSI_females <- ggplot(Data, aes(x = Condition, y = log2_HSI, fill = Condition)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_fill_manual(values = myColors) +
  geom_text(data = cld_df, aes(x = Condition, y = 3.85, label = .group),
            size = 5, color = "black", inherit.aes = FALSE) +
  geom_text(data = n_df, aes(x = Condition, y = -2.7, label = label),
            size = 4, fontface = "italic", inherit.aes = FALSE) +
  labs(y = "Log2 HSI", x = "", title = "HSI") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.text.y = element_text(size = 11),
        axis.title.y = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"))
plot_HSI_females

ggsave("HSI_plot.png", plot = plot_HSI_females,
       width = 7, height = 5, units = "in", dpi = 600)
```
## K analysis (condition factor)
```{r K}
#Diagnostic on standard lm
model_K <- lm(log2_K ~ Condition + Exposure, data = Data)
par(mfrow = c(2, 2))
plot(model_K, main = "K")
#residuals vs fitted = random scatter around zero --> good
#QQ plot some outliers --> rlm

cooksd <- cooks.distance(model_K)
threshold <- 4 / (nrow(Data) - length(coef(model_K)))
outliers <- which(cooksd > threshold)
outliers
Data[outliers, ]

#Robust regression
model_K_robust <- rlm(log2_K ~ Condition + Exposure, data = Data)

#Omnibus F-test
f.robftest(model_K_robust, var = c("ConditionAcid", "ConditionTemp", "ConditionPFOS",
                                   "ConditionAcidTemp", "ConditionAcidPFOS",
                                   "ConditionPFOSTemp", "ConditionAcidPFOSTemp"))
f.robftest(model_K_robust, var = "Exposure3")
#Huge exposure effect --> makes sense as expo 2 fish are older --> bigger

#Sensitivity check: ANOVA without outliers (exploratory)
Data_no_out <- Data[!(rownames(Data) %in% outliers), ]
model_K_aov <- aov(log2_K ~ Condition + Exposure, data = Data_no_out)
summary(model_K_aov)

#Dunnett: each treatment vs Control
emm <- emmeans(model_K_robust, ~ Condition)
contrast(emm, method = "trt.vs.ctrl", ref = "Control")

#All pairwise comparisons (BH-adjusted)
pairs(emm, adjust = "BH")

#Compact letter display
cld_df <- as.data.frame(cld(emm, Letters = letters, adjust = "BH"))
cld_df$.group <- trimws(cld_df$.group)

#Sample sizes
n_df <- Data %>%
  filter(!is.na(log2_K)) %>%
  group_by(Condition) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n))

#Plot
K_plot_females <- ggplot(Data, aes(x = Condition, y = log2_K, fill = Condition)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_fill_manual(values = myColors) +
  geom_text(data = cld_df, aes(x = Condition, y = 2.3, label = .group),
            size = 5, color = "black", inherit.aes = FALSE) +
  geom_text(data = n_df, aes(x = Condition, y = 0.1, label = label),
            size = 4, fontface = "italic", inherit.aes = FALSE) +
  labs(y = "Log2 K (condition factor)", x = "", title = "K") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.text.y = element_text(size = 11),
        axis.title.y = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"))
K_plot_females

ggsave("K_plot.png", plot = K_plot_females,
       width = 7, height = 5, units = "in", dpi = 600)
```
## Mixture effects - factorial ANOVA
The 8 conditions are a 2×2×2 factorial (Acid × Temp × PFOS presence/absence).
Decomposing Condition into 3 binary factors tests for additive vs interactive mixture effects.

```{r}
#Build 3 binary stressor factors from Condition
Data <- Data %>%
  mutate(
    Acid_f = factor(grepl("Acid", Condition), levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    Temp_f = factor(grepl("Temp", Condition), levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    PFOS_f = factor(grepl("PFOS", Condition), levels = c(FALSE, TRUE), labels = c("No", "Yes"))
  )

#Sanity check: each Condition maps to the right combination
table(Data$Condition, Data$Acid_f)
table(Data$Condition, Data$Temp_f)
table(Data$Condition, Data$PFOS_f)

#Factorial ANOVA for each endpoint
#Significant main effects --> that stressor has an effect on average across all combinations
#Significant 2-way interaction --> non-additive mixture effect (synergy or antagonism) = the effects of X differs depending on Y present or not
#Significant 3-way interaction --> triple-mixture behaves differently than 2-way predicts 

cat("\n===== LENGTH =====\n")
Anova(lm(log2_Length ~ Acid_f * Temp_f * PFOS_f + Exposure, data = Data), type ="II")

cat("\n===== WEIGHT =====\n")
Anova(lm(log2_Weight_mg ~ Acid_f * Temp_f * PFOS_f + Exposure, data = Data), type ="II")

cat("\n===== HSI =====\n")
Anova(lm(log2_HSI ~ Acid_f * Temp_f * PFOS_f + Exposure, data = Data), type ="II")

cat("\n===== K =====\n")
Anova(lm(log2_K ~ Acid_f * Temp_f * PFOS_f + Exposure, data = Data), type ="II")
```
## Create the final figure
```{r}
combined_plot <- (length_plot_females + weight_plot_females) /
                 (plot_HSI_females + K_plot_females) +
  plot_annotation(tag_levels = "A")

combined_plot

ggsave("Fig_morphology_females.png", plot = combined_plot,
       width = 14, height = 10, units = "in", dpi = 600)
```
## Summary
Robust reductions in length and weight are observed under some conditions, 
while K remains unchanged, suggesting stress-induced 
reduction in somatic growth without disproportionate changes in body composition. 
Factorial ANOVA reveals significant Acid × PFOS interactions for length and weight, 
as well as Acid × Temp and Acid × PFOS × Temp interactions for weight, indicating 
non-additive mixture effects.   

## Session info
```{r session_info}
sessionInfo()
```



